Spatio-angular fluorescence microscopy I. Basic theory (Draft: \today)

### 2018-11-21 papers/basic-theory/paper/paper.tex
New special functions, added simulation figures.

% Talon's custom commands
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\bvp}{\bs{\nu}_p}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
% \DeclareFontFamily{OT1}{pzc}{}
% \DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\author{Talon Chandler,\authormark{1,*} Authors TBD,\authormark{2,*}Rudolf Oldenbourg\authormark{2,*}, and Patrick La Rivi\`ere \authormark{1}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular dipole transfer function. We demonstrate our formalism by
  analyzing a single-view fluorescence microscope without using the monopole or
  scalar approximations. We show that this imaging system has an angular band
  limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with simple phantoms. Notably, we
  show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is recorded by fluorescence microscopes, and we show
  that these effects persist in the paraxial limit. We discuss the implications
  of our analysis for all quantitative fluorescence microscopy studies and lay
  out a path towards a complete theory.
\end{abstract*}

@@

Fluorescence microscopes are widely used in the biological sciences for
measuring the spatial distribution of fluorophores throughout a sample [cite].
While an unprocessed fluorescence micrograph reports the approximate
distribution of fluorophores throughout a sample, all microscopes are
diffraction limited [cite], so the image is a blurred version of the true
fluorophore distribution.

@@

The widely-used monopole model treats fluorophores as molecules that absorb and
emit light isotropically. Despite their use in models, monopole
absorber/emitters do not exist in nature. All real fluorophores absorb and emit
light anisotropically, and in this work we investigate the more realistic dipole
model of fluorophores. Since the absorption and emission patterns of many widely
used fluorophores including green fluroescent protein (GFP) are known to be well
described by the dipole model [cite Inoue], we start by considering the dipole
model before higher-order multipole models.

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to draw new conclusions about all fluorescence microscopes. In
Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will demonstrate the mapping between
object and data space in two different bases. In Section \ref{sec:dipole} we
will extend the monopole imaging model to dipoles and examine the mapping in
four different bases.

@@

We can represent the object as a function that assigns a real number to each
point on a plane, so we identify object space as
$\mbb{U} = \mbb{L}_2(\mbb{R}^2)$---the set of square-integrable functions on the
two-dimensional plane. Similarly, we have a two-dimensional detector that
measures a real number at every point on a plane, so data space is the same set
$\mbb{V} = \mbb{L}_2(\mbb{R}^2)$.

@@

Next, we assume that the optical system is \textit{aplanatic}---Abbe's sine
condition is satisfied and on-axis points are imaged without aberration. Abbe's
sine condition guarantees that off-axis points are imaged without spherical
aberration or coma \cite{mansuripur2009}, so the imaging system can be modeled
within the field-of-view of the optical system as a magnifier with
shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro)f(\ro), \label{eq:nonconv}
\end{align}
where $m$ is a magnification factor. Note that we have modeled the object and
data as continuous two-dimensional functions that extend to infinity. All real
objects must be restricted to a finite field-of-view and the data must be
discretely sampled, but we continue with continuous function with infinite
extent for convenience.

@@

We have chosen to normalize the monopole point spread function so that
\begin{align}
  \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}) = 1. \label{eq:norm}
\end{align}
The monopole point spread function corresponds to a
measurable irradiance, so it is always real-valued and positive.

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-block/monopole-block.pdf}
  \caption{The mapping between the object and data space of a monopole
    fluorescence microscope can be computed in two different bases---a delta
    function basis and a spatial harmonic basis. The change of basis is given by
    the two-dimensional Fourier transform denoted $\mathcal{F}_{\mbb{R}^2}$.}
     \label{fig:monopole-block}      
\end{figure}

The mapping between the object and the data in linear shift-invariant
microscopes takes a particularly simple form in a spatial harmonic basis. If we
apply the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bv) = H(\bv)F(\bv),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bv) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
\end{align}
the \textit{monopole transfer function} as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bv) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv).
\end{align}
The monopole point spread function is normalized and real-valued, so we know
that the monopole transfer function is normalized $H(0) = 1$ and
conjugate-symmetric $H(-\bv) = H^*(\bv)$ where $z^*$ denotes the complex
conjugate of $z$.

@@

Note that we have used the term \textit{monopole transfer function} instead of
the commonly-used term \textit{optical transfer function}. We reserve the term
\textit{optical transfer function} for optical systems---the optical transfer
function maps between an input irradiance spectrum and an output irradiance
spectrum in an optical system. We can use optical transfer functions to model
the optical fields in a microscope, but ultimately we are always interested in
measuring an object. Although the monopole transfer function can be identical to
the optical transfer function, we will find the distinction to be valuable when
we consider dipole transfer functions.

@@

Since the monopole transfer function is given by the two-dimensional Fourier
transform of the monopole point spread function
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\,h(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bv],
\end{align}
we can plug in Eq. \ref{eq:absquarescalar} and use the autocorrelation theorem to
rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\bvp\, C(\bvp)C^*(\bvp - \bv), 
\end{align}
where we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function
\begin{align}
  C(\bvp) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bvp].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.
We have chosen the symbol $\bvp$ as the coordinate for the coherent
transfer function because the Fourier plane of the detector is often called the
\textit{pupil plane}.

The coherent transfer function provides a valuable shortcut for analyzing
microscopes since it is usually straightforward to calculate the field in the
pupil plane. A typical approach for calculating the transfer functions is to (1)
calculate the field in the pupil plane, (2) scale the field to find the
coherent monopole transfer function, then (3) use the relationships in Figure
\ref{fig:monopole-transfer-functions} to calculate the other transfer functions.

@@

We define the same demagnified detector coordinate $\rd = \rd'/m$ and a new
normalization factor that corresponds to the total power incident on the
detector due to a spatial point source with an angularly uniform distribution of
dipoles
$P_\text{dip} = \int_{\mbb{S}^2}d\mh{s}\int_{\mbb{R}^2}d\mb{r}\, h'(m\mb{r},
\so)$. We use these scaling factors to define the
\textit{dipole point spread function} as
\begin{align}
  h(\rd - \ro, \so) = \frac{h'(m[\rd - \ro], \so)}{P_\text{dip}}, 
\end{align}
and the \textit{scaled irradiance} as 
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_\text{dip}}. 
\end{align}
With these definitions we can express the mapping between the object and the
data as
\begin{align}
g(\rd{}) = \int_{\mbb{S}^2}d\so{}\int_{\mbb{R}^2}d\ro{}\, h(\rd{} -\ro{}, \so{})f(\ro, \so). \label{eq:odpsf}
\end{align}

Similar to the monopole case, we have chosen to normalize the dipole point
spread function so that
\begin{align}
  \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \so) = 1. 
\end{align}
The dipole point spread function is a measurable quantity, so it is real-valued
and positive.

\subsubsection{Spatial dipole transfer function}
We can make our first change of basis by applying the Fourier-convolution
theorem to Eq. \ref{eq:odpsf} which yields
\begin{align}
G(\bv) = \int_{\mbb{S}^2}d\so\, H(\bv, \so)F(\bv, \so) \label{eq:odotf},
\end{align}
where we define the \textit{spatial dipole transfer function} as
  \begin{align}
  H(\bv, \so) &= \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \so)\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
  \end{align}
  and the \textit{spatial dipole spectrum} as
  \begin{align}
  F(\bv, \so) &= \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r}, \so)\, \text{exp}(-2\pi i\mb{r}\cdot\bv). 
  \end{align}
  Since the dipole point spread function is normalized and real, we know that
  the spatial dipole transfer function is normalized
  $\int_{\mbb{S}^2}d\mh{s}\, H(0, \so) = 1$ and conjugate-symmetric
  $H(-\bv, \so) = H^*(\bv, \so)$.

  This choice of basis is well suited for simulating and analyzing objects that
  are angularly sparse and spatially dense.

  \begin{figure}
  \hspace{-5em}
  \includegraphics[scale=1.0]{../figures/dipole-block/dipole-block.pdf}
  \caption{The mapping between the object and data space of a dipole
    fluorescence microscope can be computed in four different bases---a delta
    function basis, a spatial-harmonic/angular-delta basis, a
    spatial-delta/spherical-harmonic basis, and a spatio-angular harmonic basis.
    The changes of basis are given by the two-dimensional Fourier transform
    denoted $\mathcal{F}_{\mbb{R}^2}$ and the spherical Fourier transform
    denoted $\mathcal{F}_{\mbb{S}^2}$.}
   \label{fig:dipole-block}      
    \end{figure}

\subsubsection{Angular dipole transfer function}
The spherical harmonics are another set of convenient basis functions. We can
change basis from spherical delta functions to spherical harmonics by applying
the generalized Plancharel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}\, p(\mh{s})q(\mh{s}) = \lmsum P_\ell^m Q_\ell^m, \label{eq:plan}
\end{align}
where $p(\mh{s})$ and $q(\mh{s})$ are arbitrary functions on the sphere, $P_\ell^m$ and $Q_\ell^m$ are their spherical Fourier transforms defined by
\begin{align}
  P_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, p(\mh{s})Y_\ell^{m*}(\mh{s}),
\end{align}
and $Y_{\ell}^m(\mh{s})$ are the spherical harmonic functions defined in
Appendix \ref{sec:sph}. Eq. \ref{eq:plan} expresses the fact that scalar
products are invariant under a change of basis (see Eq. 3.78 of
\cite{barrett2004}). The left-hand side of Eq. \ref{eq:plan} is the scalar
product of $\mbb{L}_2(\mbb{S}^2)$ functions in a delta function basis and the
right-hand side is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a
spherical harmonic function basis. Applying Eq. \ref{eq:plan} to Eq.
\ref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum H_\ell^m(\rd - \ro)F_\ell^m(\ro), \label{eq:atf-form}
\end{align}
where we have defined the \textit{angular dipole transfer function} as
\begin{align}
  H_\ell^m(\rd - \ro) = \int_{\mbb{S}^2}d\so\, h(\rd - \ro, \so)Y_{\ell}^{m*}(\so),\label{eq:atf-prep} 
\end{align}
and the \textit{angular dipole spectrum} as
\begin{align}
  F_\ell^m(\ro) = \int_{\mbb{S}^2}d\so\, f(\ro, \so)Y_{\ell}^{m*}(\so).
\end{align}
Since the dipole point spread function is normalized and real-valued, we know
that the angular dipole transfer function is normalized
$\int_{\mbb{R}^2}d\mb{r}\, H_0^0(\mb{r}) = 1$ and conjugate symmetric
$H_\ell^{-m}(\mb{r}) = (-1)^mH_\ell^{m*}(\mb{r})$.

This basis is well suited for simulating and analyzing objects that are
spatially sparse and angularly dense.

\subsubsection{Spatio-angular dipole transfer function}
We can arrive at our final basis in two ways: by applying the generalized
Plancharel theorem for spherical functions to Eq. \ref{eq:odotf} or by applying
the Fourier convolution theorem to Eq. \ref{eq:atf-form}. We follow the
second path and find that
\begin{align}
G(\bv) = \lmsum \msf{H}_\ell^m(\bv)\msf{F}_\ell^m(\bv) \label{eq:saft},
\end{align}
where we have defined the \textit{spatio-angular dipole transfer function} as
  \begin{align}
  \msf{H}_\ell^m(\bv) &= \int_{\mbb{S}^2}d\so\, H(\bv, \so)Y_\ell^{m*}(\so),
  \end{align}
  and the \textit{spatio-angular dipole spectrum} as
  \begin{align}
  \msf{F}_\ell^m(\bv) &= \int_{\mbb{S}^2}d\so\, F(\bv, \so)Y_\ell^{m*}(\so).
  \end{align}
  Since the dipole point spread function is normalized and real-valued, we know
  that the spatio-angular dipole transfer function is normalized
  $\msf{H}_0^0(0) = 1$ and conjugate symmetric
  $\msf{H}_\ell^{-m}(-\bv) = (-1)^m\msf{H}_\ell^{m*}(\bv)$.

  This basis is well suited for simulating and analyzing arbitrary samples
  because it exploits the sparsity of the imaging system. We will see this
  advantage explicitly in Section XXX when we calculate a specific
  spatio-angular dipole transfer function and see that it has an angular band
  limit.

  Figure \ref{fig:dipole-block} summarizes the relationship between the four
  bases that we can use to compute the image of a field of dipoles. We reiterate
  that all four bases may be useful depending on the sparsity of the sample.

@@

TODO: Cite connection between dipole pupil and Jones pupil. We can use the
Jones/dipole pupil as a way to extend our model in a straightforward way. Add
phase aberration (like the Gibson-Lanni model), polarization aberrations,
amplitude masks, phase masks, etc. Also cite Piestun Green's tensor engineering.

>>>>

% Talon's custom commands
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\taup}{\bs{\tau}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
% \DeclareFontFamily{OT1}{pzc}{}
% \DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\author{Talon Chandler,\authormark{1,*} Authors TBD,\authormark{2,*}Hari Shroff\authormark{2,*}, Rudolf Oldenbourg\authormark{2,*}, and Patrick La Rivi\`ere \authormark{1}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  dipole spatio-angular transfer function. We demonstrate our formalism by
  analyzing a single-view fluorescence microscope without using the monopole or
  scalar approximations. We show that this imaging system has an angular band
  limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with simple phantoms. Notably, we
  show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is recorded by fluorescence microscopes. We discuss the
  implications of our analysis for all quantitative fluorescence microscopy
  studies and lay out a path towards a complete theory.
\end{abstract*}

@@

Fluorescence microscopes are widely used in the biological sciences for
measuring the spatial distribution of fluorophores throughout a sample [cite].
While an unprocessed fluorescence micrograph reports the approximate
distribution of fluorophores throughout a sample, all microscopes are
diffraction limited, so the image is a blurred version of the true
fluorophore distribution.

@@

The widely-used monopole model treats fluorophores as molecules that absorb and
emit light isotropically. Despite their use in models, monopole
absorbers/emitters do not exist in nature. All real fluorophores absorb and emit
light anisotropically, and in this work we investigate the more realistic dipole
model of fluorophores. Since the absorption and emission patterns of many widely
used fluorophores including green fluroescent protein (GFP) are known to be well
described by the dipole model [cite Inoue], we start by considering the dipole
model before higher-order multipole models.

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to unify the analysis of spatio-angular fluorescence microscopes.
In Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will demonstrate the mapping between
object and data space in two different bases. In Section \ref{sec:dipole} we
will extend the monopole imaging model to dipoles and examine the mapping in
four different bases.

@@

We can represent the object as a function that assigns a real number to each
point on a plane, so we identify object space as
$\mbb{U} = \mbb{L}_2(\mbb{R}^2)$---the set of square-integrable functions on the
two-dimensional plane. Similarly, we have a two-dimensional detector that
measures a real number at each point on a plane, so data space is the same set
$\mbb{V} = \mbb{L}_2(\mbb{R}^2)$.

@@

Next, we assume that the optical system is \textit{aplanatic}---Abbe's sine
condition is satisfied and on-axis points are imaged without aberration. Abbe's
sine condition guarantees that off-axis points are imaged without spherical
aberration or coma \cite{mansuripur2009}, so the imaging system can be modeled
within the field-of-view of the optical system as a magnifier with
shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro)f(\ro), \label{eq:nonconv}
\end{align}
where $m$ is a magnification factor.

@@

We have chosen to normalize the monopole point spread function so that
\begin{align}
  \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}) = 1. \label{eq:norm}
\end{align}
The monopole point spread function corresponds to a measurable irradiance, so it
is always real and positive.

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-block/monopole-block.pdf}
  \caption{The mapping between the object and data space of a monopole
    fluorescence microscope can be computed in two different bases---a delta
    function basis and a spatial harmonic basis. The change of basis can be
    computed with a two-dimensional Fourier transform denoted
    $\mathcal{F}_{\mbb{R}^2}$.}
     \label{fig:monopole-block}      
\end{figure}

The mapping between the object and the data in a linear shift-invariant imaging
system takes a particularly simple form in a spatial harmonic basis. If we apply
the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bv) = H(\bv)F(\bv),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bv) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
\end{align}
the \textit{monopole transfer function} as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bv) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv).
\end{align}
The monopole point spread function is normalized and real, so we know
that the monopole transfer function is normalized $H(0) = 1$ and
conjugate symmetric $H(-\bv) = H^*(\bv)$ where $z^*$ denotes the complex
conjugate of $z$.

@@

We have been careful to use the term \textit{monopole transfer function} instead
of the commonly-used term \textit{optical transfer function}. We reserve the
term \textit{optical transfer function} for optical systems---the optical
transfer function maps between an input irradiance spectrum and an output
irradiance spectrum in an optical system. We can use optical transfer functions
to model the propagation of light through a microscope, but ultimately we are
always interested in the object, not the light emitted by the object. We will
find the distinction between the optical transfer function and the object
transfer function to be especially valuable when we consider dipoles later in
this section.

@@

Since the monopole transfer function is given by the two-dimensional Fourier
transform of the monopole point spread function
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\,h(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bv],
\end{align}
we can plug in Eq. \ref{eq:absquarescalar} and use the autocorrelation theorem to
rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\taup\, C(\taup)C^*(\taup - \bv), 
\end{align}
where we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function
\begin{align}
  C(\taup) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.
% We have chosen the symbol $\taup$ as the coordinate for the coherent
% transfer function because the Fourier plane of the detector is often called the
% \textit{pupil plane}.

The coherent transfer function provides a valuable shortcut for analyzing
microscopes since it is often straightforward to calculate the field in a
Fourier plane of the detector. A typical approach for calculating the transfer
functions is to (1) calculate the field in a Fourier plane of the detector, (2)
scale the field to find the monopole coherent transfer function, then (3) use
the relationships in Figure \ref{fig:monopole-transfer-functions} to calculate
the other transfer functions.

@@

We define the same demagnified detector coordinate $\rd = \rd'/m$ and a new
normalization factor that corresponds to the total power incident on the
detector due to a spatial point source with an angularly uniform distribution of
dipoles
$P_\text{dip} = \int_{\mbb{S}^2}d\mh{s}\int_{\mbb{R}^2}d\mb{r}\, h'(m\mb{r},
\so)$. We use these scaling factors to define the
\textit{dipole point spread function} as
\begin{align}
  h(\rd - \ro, \so) = \frac{h'(m[\rd - \ro], \so)}{P_\text{dip}}, 
\end{align}
and the \textit{scaled irradiance} as 
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_\text{dip}}. 
\end{align}
With these definitions we can express the mapping between the object and the
data as
\begin{align}
g(\rd{}) = \int_{\mbb{S}^2}d\so{}\int_{\mbb{R}^2}d\ro{}\, h(\rd{} -\ro{}, \so{})f(\ro, \so). \label{eq:odpsf}
\end{align}
Eq. \ref{eq:odpsf} is a key result because it represents the mapping between
object space and data space in a delta function basis. The integrals in
\ref{eq:odpsf} would be extremely expensive to compute for an arbitrary object,
but the integrals simplify to an efficient sum if the object consists of
spatially fixed dipoles in a fixed orientation. In other words, Eq.
\ref{eq:odpsf} is ideal for simulating and analyzing objects that are spatially
and angularly sparse; e.g. single fluorophores that are rigidly attached to a
structure .

Similar to the monopole case, we have chosen to normalize the dipole point
spread function so that
\begin{align}
  \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \so) = 1. 
\end{align}
The dipole point spread function is a measurable quantity, so it is real
and positive.

\subsubsection{Spatial dipole transfer function}
We can make our first change of basis by applying the Fourier-convolution
theorem to Eq. \ref{eq:odpsf} which yields
\begin{align}
G(\bv) = \int_{\mbb{S}^2}d\so\, H(\bv, \so)F(\bv, \so) \label{eq:odotf},
\end{align}
where we define the \textit{spatial dipole transfer function} as
  \begin{align}
  H(\bv, \so) &= \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \so)\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
  \end{align}
  and the \textit{spatial dipole spectrum} as
  \begin{align}
  F(\bv, \so) &= \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r}, \so)\, \text{exp}(-2\pi i\mb{r}\cdot\bv). 
  \end{align}
  Since the dipole point spread function is normalized and real, we know that
  the spatial dipole transfer function is normalized
  $\int_{\mbb{S}^2}d\mh{s}\, H(0, \so) = 1$ and conjugate symmetric
  $H(-\bv, \so) = H^*(\bv, \so)$.

  This basis is ideal for simulating and analyzing objects that are angularly
  sparse and spatially dense; e.g. rod structures that contain fluorophores in a
  fixed orientation.

  \begin{figure}
  \hspace{-5em}
  \includegraphics[scale=1.0]{../figures/dipole-block/dipole-block.pdf}
  \caption{The mapping between the object and data space of a dipole imaging
    system can be computed in four different bases---a delta function basis, a
    spatial-harmonic/angular-delta basis, a spatial-delta/spherical-harmonic
    basis, and a spatio-angular harmonic basis. The changes of basis can be
    computed with the two-dimensional Fourier transform denoted
    $\mathcal{F}_{\mbb{R}^2}$, and the spherical Fourier transform denoted
    $\mathcal{F}_{\mbb{S}^2}$.}
   \label{fig:dipole-block}      
    \end{figure}

\subsubsection{Dipole angular transfer function}
The spherical harmonics are another set of convenient basis functions that are
play the same role as complex exponentials in spatial transfer functions---see
Appendix \ref{sec:sph} for an introduction to the spherical harmonics. We can
change basis from spherical delta functions to spherical harmonics by applying
the generalized Plancharel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}\, p(\mh{s})q^*(\mh{s}) = \lmsum P_\ell^m Q_\ell^{m*}, \label{eq:plan}
\end{align}
where $p(\mh{s})$ and $q(\mh{s})$ are arbitrary functions on the sphere,
$P_\ell^m$ and $Q_\ell^m$ are their spherical Fourier transforms defined by
\begin{align}
  P_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, p(\mh{s})Y_\ell^{m*}(\mh{s}),
\end{align}
and $Y_{\ell}^m(\mh{s})$ are the spherical harmonic functions defined in
Appendix \ref{sec:sph}. Eq. \ref{eq:plan} expresses the fact that scalar
products are invariant under a change of basis (see Eq. 3.78 of
\cite{barrett2004}). The left-hand side of Eq. \ref{eq:plan} is the scalar
product of $\mbb{L}_2(\mbb{S}^2)$ functions in a delta function basis and the
right-hand side is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a
spherical harmonic function basis. Applying Eq. \ref{eq:plan} to Eq.
\ref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum H_\ell^m(\rd - \ro)F_\ell^m(\ro), \label{eq:atf-form}
\end{align}
where we have defined the \textit{dipole angular transfer function} as
\begin{align}
  H_\ell^m(\rd - \ro) = \int_{\mbb{S}^2}d\so\, h(\rd - \ro, \so)Y_{\ell}^{m*}(\so),\label{eq:atf-prep} 
\end{align}
and the \textit{dipole angular spectrum} as
\begin{align}
  F_\ell^m(\ro) = \int_{\mbb{S}^2}d\so\, f(\ro, \so)Y_{\ell}^{m*}(\so).
\end{align}
Since the dipole point spread function is normalized and real, we know
that the dipole angular transfer function is normalized
$\int_{\mbb{R}^2}d\mb{r}\, H_0^0(\mb{r}) = 1$ and conjugate symmetric
$H_\ell^{-m}(\mb{r}) = (-1)^mH_\ell^{m*}(\mb{r})$.

This basis is well suited for simulating and analyzing objects that are
spatially sparse and angularly dense; e.g. rotating single fluorophores that are
spatially fixed, or many fluorophores that are close together with varying
orientations.

\subsubsection{Spatio-angular dipole transfer function}
We can arrive at our final basis in two ways: by applying the generalized
Plancharel theorem for spherical functions to Eq. \ref{eq:odotf} or by applying
the Fourier convolution theorem to Eq. \ref{eq:atf-form}. We follow the
second path and find that
\begin{align}
G(\bv) = \lmsum \msf{H}_\ell^m(\bv)\msf{F}_\ell^m(\bv) \label{eq:saft},
\end{align}
where we have defined the \textit{dipole spatio-angular transfer function} as
  \begin{align}
  \msf{H}_\ell^m(\bv) &= \int_{\mbb{S}^2}d\so\, H(\bv, \so)Y_\ell^{m*}(\so),
  \end{align}
  and the \textit{dipole spatio-angular spectrum} as
  \begin{align}
  \msf{F}_\ell^m(\bv) &= \int_{\mbb{S}^2}d\so\, F(\bv, \so)Y_\ell^{m*}(\so).
  \end{align}
  Since the dipole point spread function is normalized and real, we know
  that the dipole spatio-angular transfer function is normalized
  $\msf{H}_0^0(0) = 1$ and conjugate symmetric
  $\msf{H}_\ell^{-m}(-\bv) = (-1)^m\msf{H}_\ell^{m*}(\bv)$.

  This basis is well suited for simulating and analyzing arbitrary samples
  because it exploits the band limit of the imaging system. We will see this
  advantage explicitly in Section \ref{sec:results} when we calculate a specific
  dipole spatio-angular transfer function and see that it has spatial and
  angular band limits.

  Figure \ref{fig:dipole-block} summarizes the relationship between the four
  bases that we can use to compute the image of a field of dipoles. We reiterate
  that all four bases may be useful depending on the sample.

@@

TODO: Cite connection between dipole pupil and Jones pupil. We can use the
Jones/dipole pupil as a way to extend our model in a straightforward way. Add
phase aberration (like the Gibson-Lanni model), polarization aberrations,
amplitude masks, phase masks, etc.

TODO: Cite Piestun Green's tensor engineering.
